---
title: "TCGA Kinase Mutation Analysis"
output:
  html_document:
    code_folding: hide
---

```{r library_loading}
library(FirebrowseR)
library(DarkKinaseTools)
library(BerginskiRMisc)
library(readr)
library(here)
library(tidyverse)
library(progress)
```

## Data Collection/Loading

This notebook assumes that you have already run the kinase_mutant_download.Rmd
document and successfully collected the kinase_MAF.rds file.

```{r data_collection, cache=TRUE}

#check if a data file has been made with these results, if so, go ahead and load
#it, otherwise WARNING!
if (file.exists(here('data','kinase_MAF.rds'))) {
  kinase_MAF = read_rds('data/kinase_MAF.rds')
} else {
  warning("Couldn't find the kinase_MAF.rds file, did you run the kinase_mutant_download document?")
}

```

## Data Organization/Analysis

The MAF data is organized to indicate each individual mutation call made in for
each kinase in each tumor sample. Let's get a count of how often each kinase is
mutated in each patient and summarize that as a percent.

```{r data_analysis, cache=TRUE}
#Get the number of samples for each cohort and filter down to only the tumor
#samples, i.e. not the normal paired samples
sample_counts = Metadata.Counts(format='csv') %>% filter(sample_type == "Tumor")

#Count the number of samples, by the unique tumor sample barcode, that each
#cohort, gene and variant type
kinase_mutation_count = kinase_MAF %>%
  group_by(cohort,Hugo_Symbol,Variant_Classification) %>%
  summarise(num_times_mutated = length(unique(Tumor_Sample_Barcode)))

#We'd also like to count the number of times a sample contains one of the frame
#shift mutations. I couldn't figure out a way to do this with a single dplyr
#pipe sequence, so I've setup a second count and assigned these counts to the
#Variant_Classification of simply "Frameshift"
frameshift_variants = c("Frame_Shift_Del", "In_Frame_Ins", "In_Frame_Del",
                        "Frame_Shift_Ins")

#Want to summarize the samples that are hit by a frame shift mutation, making a new 
kinase_frameshift_count = kinase_MAF %>%
  filter(Variant_Classification %in% frameshift_variants) %>%
  group_by(cohort,Hugo_Symbol) %>%
  summarise(num_times_mutated = length(unique(Tumor_Sample_Barcode))) %>%
  mutate(Variant_Classification = "Frameshift")


kinase_mutation_count = rbind(kinase_mutation_count,kinase_frameshift_count)

#bring together the counts of patient samples by cohort (I'm not sure what the
#exact difference is between the maf and rawmaf count is, but they differ) with
#the counts of how often each mutation is occuring in each kinase
kinase_mutation_count = left_join(kinase_mutation_count,
                                  sample_counts %>% select(cohort,maf,rawmaf))

kinase_mutation_count = kinase_mutation_count %>%
  mutate(percent_mutation = 100*(num_times_mutated/max(c(rawmaf,maf))))

#I can't think of an elegant way to convert the percent mutated number above
#into a summary table with freq_missensemutation, freq_nonsensemutation,
#freq_frameshiftmutation, freq_splicesitemutation, so I'll just brute force it
#with a for loops
Hugo_names = unique(kinase_mutation_count$Hugo_Symbol)
cohorts = unique(kinase_mutation_count$cohort)

kinase_mutation_freq = expand.grid(cohorts,Hugo_names);
kinase_mutation_freq = cbind(
  expand.grid(cohorts,Hugo_names),
  rep(0,length(Hugo_names)*length(cohorts)),
  rep(0,length(Hugo_names)*length(cohorts)),
  rep(0,length(Hugo_names)*length(cohorts)),
  rep(0,length(Hugo_names)*length(cohorts)),
  rep(0,length(Hugo_names)*length(cohorts)))

names(kinase_mutation_freq) <-
  c('cohort','gene_id','freq_missensemutation','freq_nonsensemutation',
    'freq_frameshiftmutation','freq_splicesitemutation',
    'freq_silentmutation')
  
this_bar = progress_bar$new(total = dim(kinase_mutation_count)[1]);
for (row_num in 1:dim(kinase_mutation_count)[1]) {
  this_data = kinase_mutation_count[row_num,]
  target_row = which(kinase_mutation_freq$cohort == this_data$cohort & 
                     kinase_mutation_freq$gene_id == this_data$Hugo_Symbol)
  if (this_data$Variant_Classification == "Missense_Mutation") {
    kinase_mutation_freq$freq_missensemutation[target_row] = this_data$percent_mutation
  } else if (this_data$Variant_Classification == "Nonsense_Mutation") {
    kinase_mutation_freq$freq_nonsensemutation[target_row] = this_data$percent_mutation
  } else if (this_data$Variant_Classification == "Frameshift") {
    kinase_mutation_freq$freq_frameshiftmutation[target_row] = this_data$percent_mutation
  } else if (this_data$Variant_Classification == "Splice_Site") {
    kinase_mutation_freq$freq_splicesitemutation[target_row] = this_data$percent_mutation
  } else if (this_data$Variant_Classification == "Silent") {
    kinase_mutation_freq$freq_silentmutation[target_row] = this_data$percent_mutation
  }
  
  this_bar$tick()
}

#Lets add the light/dark kinase class distinction
kinase_mutation_freq = left_join(kinase_mutation_freq,
                                 all_kinases %>% select(symbol,class),
                                 by=c("gene_id"="symbol"))
```

## Data Visualization {.tabset}

Let's start the data analysis with some heatmaps of the frequency of various types of mutation present in the dark kinases.

### Missense Mutations

```{r, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_missensemutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_missensemutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_missensemutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Missense Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```

### Nonsense Mutations

```{r, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_nonsensemutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_nonsensemutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_nonsensemutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Nonsense Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```

### Frameshift Mutations

```{r, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_frameshiftmutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_frameshiftmutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_frameshiftmutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Framshift Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```

### Splicesite Mutations

```{r, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_splicesitemutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_splicesitemutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_splicesitemutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Splicesite Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```

### Silent Mutations

```{r, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_silentmutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_silentmutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_silentmutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Silent Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```