---
title: "TCGA Kinase Mutation Analysis"
output:
  html_document:
    code_folding: hide
---

```{r library_loading}
library(FirebrowseR)
library(DarkKinaseTools)
library(BerginskiRMisc)
library(readr)
library(here)
library(tidyverse)
library(progress)
```

## Data Collection/Loading

This notebook assumes that you have already run the kinase_mutant_download.Rmd
document and successfully collected the kinase_MAF.rds file.

```{r data_collection, cache=TRUE}

#check if a data file has been made with these results, if so, go ahead and load
#it, otherwise WARNING!
if (file.exists(here('data','kinase_MAF.rds'))) {
  kinase_MAF = read_rds('data/kinase_MAF.rds')
} else {
  warning("Couldn't find the kinase_MAF.rds file, did you run the kinase_mutant_download document?")
}

```

## Data Organization/Analysis

The MAF data is organized to indicate each individual mutation call made in for
each kinase in each tumor sample. Let's get a count of how often each kinase is
mutated in each patient and summarize that as a percent.

```{r data_analysis, cache=TRUE}
#Get the number of samples for each cohort and filter down to only the tumor
#samples, i.e. not the normal paired samples
sample_counts = Metadata.Counts(format='csv') %>% filter(sample_type == "Tumor")

#We'd also like to count the number of times a sample contains one of the frame
#shift mutations. These variants should all be counted as frame shift mutants, I
#think
frameshift_variants = c("Frame_Shift_Del", "In_Frame_Ins", "In_Frame_Del",
                        "Frame_Shift_Ins")

#This piped dplyr call takes the raw mutation data: 
#  -counts the number of hits in various mutation classes
#  -converts those counts to percentages based on the number of samples 
#   collected by TCGA
#  -adds the light/dark classification for each kinase
#
# P.S. - Sorry this probably a bit confusing, but I couldn't help myself
kinase_mutation_freq = kinase_MAF %>%
  
  #This command fills out blank entires for each combonation of cohort and
  #Hugo_symbol, useful for the later plotting steps
  complete(cohort,Hugo_Symbol) %>%
  group_by(cohort,Hugo_Symbol) %>%
  
  #First get the counts of various types of mutations
  summarise(missensemutation_count = sum(Variant_Classification %in% "Missense_Mutation"),
            nonsensemutation_count = sum(Variant_Classification %in% "Nonsense_Mutation"),
            frameshiftmutation_count = sum(Variant_Classification %in% frameshift_variants),
            splicesitemutation_count = sum(Variant_Classification %in% "Splice_Site"),
            silentmutation_count = sum(Variant_Classification %in% "Silent")) %>%
  
  #Now get the number of samples in each tumor cohort from the TCGA data, as far
  #as I can tell this number of samples is stored as maf or rawmaf, depending on
  #the cohort study
  left_join(sample_counts %>% select(cohort,maf,rawmaf)) %>%
  
  #Convert the raw counts to frequencies of mutation
  mutate(freq_missensemutation = 100*(missensemutation_count/max(c(rawmaf,maf))),
         freq_nonsensemutation = 100*(nonsensemutation_count/max(c(rawmaf,maf))),
         freq_frameshiftmutation = 100*(frameshiftmutation_count/max(c(rawmaf,maf))),
         freq_splicesitemutation = 100*(splicesitemutation_count/max(c(rawmaf,maf))),
         freq_silentmutation = 100*(silentmutation_count/max(c(rawmaf,maf)))) %>%
  rename(gene_id = Hugo_Symbol) %>%
  
  #remove the extra unneeded count columns and sample counts
  select(-ends_with('_count'),-matches('rawmaf'),-matches('maf')) %>%
  
  #finally, using the all_kinases list from DarkKinaseTools, add in a column for
  #the type (light/dark) of each kinase
  left_join(DarkKinaseTools::all_kinases %>% select(symbol,class),
            by=c("gene_id"="symbol"))
```

## Data Visualization {.tabset}

Let's start the data analysis with some heatmaps of the frequency of various types of mutation present in the dark kinases.

### Missense Mutations

```{r missense, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_missensemutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_missensemutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_missensemutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Missense Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```

### Nonsense Mutations

```{r nonsense, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_nonsensemutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_nonsensemutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_nonsensemutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Nonsense Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```

### Frameshift Mutations

```{r frameshift, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_frameshiftmutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_frameshiftmutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_frameshiftmutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Framshift Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```

### Splicesite Mutations

```{r splice, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_splicesitemutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_splicesitemutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_splicesitemutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Splicesite Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```

### Silent Mutations

```{r silent, fig.width=17.5}
dark_mutation_freq = kinase_mutation_freq %>% filter(class=="Dark")

kinase_sort_order = dark_mutation_freq %>%
  group_by(gene_id) %>%
  summarise(average_mutation = mean(freq_silentmutation)) %>%
  arrange(desc(average_mutation))

cohort_sort_order = dark_mutation_freq %>%
  group_by(cohort) %>%
  summarise(average_mutation = mean(freq_silentmutation)) %>%
  arrange(average_mutation)

ggplot(dark_mutation_freq,aes(x=gene_id,y=cohort,fill=freq_silentmutation)) + 
  geom_tile() + 
  scale_fill_distiller(direction=1,palette = 'YlOrRd') + 
  scale_x_discrete(limits=kinase_sort_order$gene_id) +
  scale_y_discrete(limits=cohort_sort_order$cohort) +
  theme_berginski() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.title = element_text(size=25)) +
  ggtitle("Frequency of Silent Mutations in TCGA Dark Kinases") +
  labs(x='Hugo Gene Symbol',y='TCGA Cancer Cohort',fill="% Patient Samples")
```