---
title: "PTEN BRCA Split Processing"
author: "Matthew Berginski"
output: github_document
---

```{r setup, include=FALSE}
suppressPackageStartupMessages(library(FirebrowseR))
suppressPackageStartupMessages(library(DarkKinaseTools))
suppressPackageStartupMessages(library(BerginskiRMisc))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(progress))
```

## Download Expression Data

I downloaded all of the mRNA sequencing data from firehose, specifically this fantastically long filename:

http://gdac.broadinstitute.org/runs/stddata__2016_01_28/data/BRCA/20160128/gdac.broadinstitute.org_BRCA.Merge_rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__RSEM_genes__data.Level_3.2016012800.0.0.tar.gz

As far as I can tell this is the full set of RNAseq data from the breast cancer patients (BRCA). I then uncompressed it in the PTEN_analysis folder.

## Load mRNA Data and Filter

This command loads all the downloaded mRNA data into memory, it takes a few minutes to run.

```{r PTEN Expression Loading}
BRCA_mRNA = read_delim(here('PTEN_analysis/gdac.broadinstitute.org_BRCA.Merge_rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__RSEM_genes__data.Level_3.2016012800.0.0/BRCA.rnaseqv2__illuminahiseq_rnaseqv2__unc_edu__Level_3__RSEM_genes__data.data.txt'), 
                  delim ="\t")


```

## Get PTEN CNV Status

```{r}
BRCA_CNV = read_rds(here('PTEN_analysis/CNV_data/BRCA_cnv.rds'))
PTEN_CNV = BRCA_CNV[["PTEN"]]

#Build a regular expression to search through the column names for our patient
#barcodes and filter the mRNA data set down to just them. Also recast as a
#data.frame to make the tidy process easier
PTEN_CNV_delete_normal = PTEN_CNV %>%
  filter(cn_alteration == -2 | cn_alteration == 0) %>%
  summarise(str = paste0(c(tcga_participant_barcode), collapse="|"))

PTEN_CNV_delete_normal$str = paste0(PTEN_CNV_delete_normal$str,"|Hybridization REF")

BRCA_mRNA_filtered = BRCA_mRNA[,grepl(PTEN_CNV_delete_normal$str,names(BRCA_mRNA))]

#Now the mRNA file is formatted so the first entry is the type of measurement in
#that column, I'm going to filter down to just the raw count columns and the
#gene_id column and then convert the data into tidy format

BRCA_mRNA_filtered = BRCA_mRNA_filtered[,which(BRCA_mRNA_filtered[1,] == "scaled_estimate" | BRCA_mRNA_filtered[1,] == "gene_id")] %>%
  separate(`Hybridization REF`, c('gene_id',NA), sep="\\|") %>%
  #this tosses the first row, which now only contains the data types, which we
  #have already filtered down to just the column we want
  slice(-1) %>%
  filter(gene_id != "?") %>%
  gather("TCGA_barcode","scaled_estimate",-gene_id) %>%
  mutate(tcga_participant_barcode = str_extract(TCGA_barcode,'[^-]*-[^-]*-[^-]*'))

#Reset the class of our columns
BRCA_mRNA_filtered$gene_id = as.factor(BRCA_mRNA_filtered$gene_id)
BRCA_mRNA_filtered$TCGA_barcode = as.factor(BRCA_mRNA_filtered$TCGA_barcode)
BRCA_mRNA_filtered$tcga_participant_barcode = as.factor(BRCA_mRNA_filtered$tcga_participant_barcode)
# BRCA_mRNA_filtered$raw_count = as.numeric(BRCA_mRNA_filtered$raw_count)
BRCA_mRNA_filtered$scaled_estimate = as.numeric(BRCA_mRNA_filtered$scaled_estimate)


BRCA_mRNA_filtered = BRCA_mRNA_filtered %>% left_join(PTEN_CNV)
```

```{r overall diffs}
mRNA_averages = BRCA_mRNA_filtered %>% 
  group_by(gene_id,cn_alteration) %>%
  summarize(average_count = mean(scaled_estimate))

mRNA_diffs = left_join(
  mRNA_averages %>% filter(cn_alteration == 0) %>% select(-cn_alteration) %>% rename(PTEN_normal_count = average_count),
  mRNA_averages %>% filter(cn_alteration == -2) %>% select(-cn_alteration) %>% rename(PTEN_deleted_count = average_count)
) %>% 
  filter(PTEN_normal_count > 2, PTEN_deleted_count > 2) %>%
  mutate(expression_ratio = log2(PTEN_normal_count/PTEN_deleted_count),
         expression_diff = PTEN_normal_count - PTEN_deleted_count)

mRNA_diff_hits = mRNA_diffs %>%
  filter(abs(expression_ratio) > 2) %>%
  arrange(desc(abs(expression_ratio)))

BRCA_log2_diff_data = BRCA_mRNA_filtered %>%
  filter(gene_id %in% mRNA_diff_hits$gene_id[1:10]) %>%
  mutate(delete_text = ifelse(cn_alteration == 0, 'Normal','Deleted'))


```

```{r syn lethal hits}
BRCA_syn_lethal = read_rds(here('PTEN_analysis/syn_lethal_hits/BRCA_syn_lethal.rds'))
```